---
title: "Intial QC"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    code_folding: hide
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir =
  paste0(Sys.getenv("RKJCOLLAB"),
  "/Training_Programs/PRIDE/PRIDE_Metabolomics/GitHub/output")) })

---

```{r setup}
# Set working directory - change work_dir to folder that contains input data and
# has a place for output data
work_dir <- paste0(
  Sys.getenv("RKJCOLLAB"), "/Training_Programs/PRIDE/PRIDE_Metabolomics/GitHub")
knitr::opts_knit$set(root.dir = work_dir)

# Load libraries
# If need to install tidyverse or PCAtools, uncomment the lines below
# install.packages("tidyverse")
library(tidyverse)
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("PCAtools")# when asked to update packages, say "no"
library(PCAtools)

```


```{r load_data}
# Load input data
intensities <- read.delim("input/compound_intensities_init.txt", stringsAsFactors = F)
sample.info <- read.delim("input/sample_info.txt", stringsAsFactors = F)
compound.info <- read.delim("input/compound_info.txt", stringsAsFactors = F)
  # compound.info not used anywhere but let's understand the information in here
```



## Data exploration

### Initial inspection of input

```{r inspect}
dim(intensities)
head(intensities[,1:10])
dim(sample.info)
head(sample.info)
head(compound.info)
```

### Initial number of samples and compounds

```{r n_m}
table(sample.info$class)
paste0("Number compounds: ", dim(intensities)[2]-1)
```


### Sample QC

#### PCA Scree plot

```{r scree}
ordered.intensities <-
  merge(sample.info[,c("sample_id", "injection_order")], intensities)
# Order by injection order - then the order of the samples will be the same as in sample.info
ordered.intensities <- ordered.intensities[order(ordered.intensities$injection_order),]
sample.ids <- ordered.intensities$sample_id
pca.mat <- as.matrix(t(ordered.intensities[,-c(1,2)]))
# Remove compounds with no variability so PCA does not fall over
no.var.compounds <- which(apply(pca.mat, 1, var) == 0)
if (length(no.var.compounds) > 0) {
  pca.mat <- pca.mat[-no.var.compounds,]  
}
# The colnames of the PCA input matrix and the rownames of the PCA metadata must
# be set and must contain the same IDs and in the same order
colnames(pca.mat) <- sample.ids
rownames(sample.info) <- sample.info$sample_id
pca = pca(pca.mat, metadata = sample.info, center=TRUE, scale=TRUE)
screeplot(pca, components=getComponents(pca, 1:20), hline = 50)
```

#### PCA scatterplots

The dashed horizontal and vertical lines denote the specificied number of SD
away from the mean, to be used for identifying outliers

##### To change

* Based on the scree plot above, modify the below if necessary to include lower
  order principal component plots (by setting the x and y parameters of the
  biplot object)
* Change the nr.sd variable to the number of SD from the mean that should be
  used as threshold to identify outlier samples

```{r pca}
nr.sd <- 3
pca.frame <- pca$rotated
pca.frame$batch <- sample.info$batch
pca.frame$class <- sample.info$class

# PC1 and PC2
x.min <- mean(pca.frame$PC1) - nr.sd*sd(pca.frame$PC1)
x.max <- mean(pca.frame$PC1) + nr.sd*sd(pca.frame$PC1)
y.min <- mean(pca.frame$PC2) - nr.sd*sd(pca.frame$PC2)
y.max <- mean(pca.frame$PC2) + nr.sd*sd(pca.frame$PC2)
ggplot(pca.frame, aes(x=PC1, y=PC2, shape=batch, color=class)) +
  geom_point() +
  scale_shape_manual(values=1:8) +
  geom_hline(yintercept=y.min, linetype="dashed") +
  geom_hline(yintercept=y.max, linetype="dashed") +
  geom_vline(xintercept=x.min, linetype="dashed") +
  geom_vline(xintercept=x.max, linetype="dashed") 

```

#### Remove PCA outliers

Remove samples that are PCA outliers (example code that is commented out)

##### To change

Change the below according to the thresholds decided from the PCA scatterplots
above

```{r pca_outliers}
# #PC1 
# PC1.min <- mean(pca.frame$PC1) - nr.sd*sd(pca.frame$PC1)
# PC1.max <- mean(pca.frame$PC1) + nr.sd*sd(pca.frame$PC1)
# PC1.outliers <- sample.info$sample_id[(pca.frame$PC1 < PC1.min) | (pca.frame$PC1 > PC1.max)]
# 
# #PC2
# PC2.min <- mean(pca.frame$PC2) - nr.sd*sd(pca.frame$PC2)
# PC2.max <- mean(pca.frame$PC2) + nr.sd*sd(pca.frame$PC2)
# PC2.outliers <- sample.info$sample_id[(pca.frame$PC2 < PC2.min) | (pca.frame$PC2 > PC2.max)]
# 
# #Create output data frame with outliers
# outlier.frame <- data.frame(
#   sample_id=c(PC1.outliers, PC2.outliers),
#   reason=c(rep("PC1 outlier", length(PC1.outliers)), rep("PC2 outlier", length(PC2.outliers)))
# )
# dim(outlier.frame)
# knitr::kable(outlier.frame)
# outlier.ids <- unique(outlier.frame$sample_id)
# length(outlier.ids)
# 
# #Remove the outliers from the intensities data frame, so they are no longer present for the 
# #subsequent QC of the comounds
# dim(intensities)
# intensities <- intensities[!(intensities$sample_id %in% outlier.ids),]
# dim(intensities)
# dim(sample.info)
# sample.info <- sample.info[!(sample.info$sample_id %in% outlier.ids),]
# dim(sample.info)
```

#### Number of peaks per sample ordered by injection order

```{r nr_peaks}
nrPeaks <- function(i) {
  return (sum(i != 1))  # in this data set, 1 denotes a missing value
}
ordered.intensities <- merge(sample.info[,c("sample_id", "injection_order")], intensities)
# Order by injection order - then the order of the samples will be the same as
# in sample.info
ordered.intensities <- ordered.intensities[order(ordered.intensities$injection_order),-c(1,2)]
nr.peaks <- apply(ordered.intensities, 1, nrPeaks)
nr.peaks.frame <- sample.info[order(sample.info$injection_order),]
nr.peaks.frame$nr_peaks <- nr.peaks
ggplot(nr.peaks.frame, aes(x=injection_order, y=nr_peaks, shape=batch, color=class)) +
  geom_point() +
  scale_shape_manual(values=1:8) 
```

#### Mean intensity per sample ordered by injection order

```{r mean_intensity}
meanIntensity <- function(i) {
  return (mean(i[i!=1]))
}
mean.intensity <- apply(ordered.intensities, 1, meanIntensity)
avg.intensity.frame <- sample.info[order(sample.info$injection_order),]
avg.intensity.frame$mean_intensity <- mean.intensity
ggplot(avg.intensity.frame,
       aes(x=injection_order, y=mean_intensity, shape=batch, color=class)) +
  geom_point() +
  scale_shape_manual(values=1:8)
```

#### Intensity boxplots

```{r boxplots}
wide.frame <- merge(intensities, sample.info)
keycol <- "compound_id"
valuecol <- "intensity"
gathercols <- names(wide.frame)[grep("^C", names(wide.frame))]
long.frame <- pivot_longer(
  wide.frame, cols = all_of(gathercols), names_to = keycol, values_to = valuecol)
long.frame <- long.frame[order(long.frame$injection_order),]
long.frame$log2_intensity <- log2(long.frame$intensity)
long.frame <- long.frame[long.frame$log2_intensity > 0,]  # don't want to plot missing values
long.frame$day_order <- NA

# For each batch, subtract the start_injection_order-1, so that the injection
# order all starts with 1 in each day
for (batch in unique(long.frame$batch)) {
  min.order <- min(long.frame$injection_order[long.frame$batch == batch])-1
  long.frame$day_order[long.frame$batch == batch] <-
    long.frame$injection_order[long.frame$batch == batch] - min.order
}
long.frame$day_order <- as.factor(long.frame$day_order)
ggplot(long.frame, aes(x = day_order, y = log2_intensity, color=class)) +
         geom_boxplot() +
        facet_wrap( ~ batch)
ggplot(long.frame, aes(x = class, y = log2_intensity, color=class)) +
         geom_boxplot() +
        facet_wrap( ~ batch) +
        theme(axis.title.x=element_blank(),
              axis.text.x=element_blank(),
              axis.ticks.x=element_blank())
```

### Compound QC

#### Intensity boxplots

```{r compound intensity}
boxplot(log2_intensity ~ batch, data=long.frame)
```

#### Histogram of percent missing values per compound

##### To change

* The line that sets compound.cols may need to change according to the naming
  convention in the compound intensities file; the current code assumes that
  they all start with C.

```{r distro_missing_compounds}
compound.frame <- merge(sample.info, intensities)
compound.cols <- names(compound.frame)[startsWith(names(compound.frame), "C")]
percMissCompounds <- function(i) {
  return ((sum(i==1)/length(i))*100)
}
compound.miss.frame <- compound.frame %>% 
  group_by(class) %>%
  summarize_at(vars(all_of(compound.cols)), list(percMissCompounds))
head(compound.miss.frame[,1:10])
keycol <- "compound_id"
valuecol <- "percent_missing"
compound.miss.long.frame <- pivot_longer(
  compound.miss.frame, cols = all_of(compound.cols), names_to = keycol,
  values_to = valuecol)
ggplot(compound.miss.long.frame, aes(x=percent_missing, fill=class)) +
    geom_histogram(position = 'dodge', breaks=seq(0, 100, 10)) +
    scale_x_continuous(breaks = seq(0, 100, 10))
```

#### Identify compounds with large missingness in research samples

##### To change

* Set the qc.classes variable to the possible class values of QC samples
* Set the max.perc.research.missingness variable to the maximum percentage
  allowed missingness in research samples; defaults to 20\%

```{r remove_high_missing_compounds}
qc.classes <- c("instrument_QC", "prep_QC")
max.perc.research.missingness <- 20
compound.research.frame <-
  merge(sample.info[!(sample.info$class %in% qc.classes),], intensities)
compound.cols <-
  names(compound.research.frame)[startsWith(names(compound.research.frame), "C")]
compound.research.miss.frame <- compound.research.frame %>% 
  summarize_at(vars(all_of(compound.cols)), list(percMissCompounds))
head(compound.research.miss.frame[,1:10])
head(t(compound.research.miss.frame[,1:10]))
compound.research.miss.long.frame <- t(compound.research.miss.frame)
sum(compound.research.miss.long.frame > max.perc.research.missingness)
miss.remove.list <-
  rownames(compound.research.miss.long.frame)[compound.research.miss.long.frame >
                                                max.perc.research.missingness]
```

#### Identify compounds present only in research samples

These compounds must be removed as they cannot be normalized by methods that
assume QC samples.
Note that if a compound has a large % missingness in QC samples (but is detected
in a small number of QC samples]), it is indicative of perhaps not really being
detected in the QC samples, and hence these compounds should be removed too 
(too much missingness in QC samples may also affect normalization).

##### To change

* Set the max.perc.qc.missingness variable to the maximum percentage allowed
  missingness in QC samples; defaults to 80\%

```{r remove_research_only_compounds}
max.perc.qc.missingness <- 80
compound.qc.miss.long.frame <-
  compound.miss.long.frame[compound.miss.long.frame$class %in% qc.classes,]
compound.qc.miss.long.frame <- compound.qc.miss.long.frame[
  !(compound.qc.miss.long.frame$compound_id %in% miss.remove.list),]
researchonly.remove.list <-
  unique(compound.qc.miss.long.frame$compound_id[
    compound.qc.miss.long.frame$percent_missing > max.perc.qc.missingness])
length(researchonly.remove.list)
```

#### Identify compounds with large CV in QC samples

##### To change

* Set the cv.threshold variable to the maximum CV allowed in QC samples;
  defaults to 0.3

```{r remove_CV_compounds}
cv.threshold <- 0.3
compound.qc.frame <- merge(
  sample.info[(sample.info$class %in% qc.classes),],
  intensities[,!(colnames(intensities) %in%
                   c(miss.remove.list, researchonly.remove.list))])
compound.cols <-
  names(compound.qc.frame)[startsWith(names(compound.qc.frame), "C")]
cV <- function(x) {
  x <- x[x != 1]
  return(sd(x)/mean(x))
}
cv.frame <- compound.qc.frame %>% 
  group_by(class) %>%
  summarize_at(vars(all_of(compound.cols)), list(cV))
head(cv.frame[,1:10])
keycol <- "compound_id"
valuecol <- "cv"
cv.long.frame <- pivot_longer(
  cv.frame, cols = all_of(compound.cols), names_to = keycol, values_to = valuecol)
ggplot(cv.long.frame, aes(x=cv, fill=class)) +
    geom_histogram(position = 'dodge',
                   breaks=seq(0, max(cv.long.frame$cv)+0.1, 0.1)) +
    scale_x_continuous(breaks = seq(0, max(cv.long.frame$cv+0.1), 0.1))
cv.remove.list <- unique(
  cv.long.frame$compound_id[cv.long.frame$cv > cv.threshold])
length(cv.remove.list)

```

#### Write sample and compound failures and create a clean output file

```{r write_output}
# Write files with information on sample and compound failures
# write.table(outlier.frame,
#             "output/sample_QC_failures.txt",
#             sep="\t", quote=F, row.names=F, col.names=T)
compound.failure.frame <- data.frame(
  compound_id=c(miss.remove.list, researchonly.remove.list, cv.remove.list),
  reason=c(rep("High % missing values", length(miss.remove.list)),
           rep("Not in QC samples", length(researchonly.remove.list)),
           rep("High CV in QC samples", length(cv.remove.list))))
write.table(compound.failure.frame,
            "output/compound_QC_failures.txt",
            sep="\t", quote=F, row.names=F, col.names=T)

# Write clean files with sample and compound QC failures removed
# (note that sample failures were removed from intensities prior to compound QC)
keep.cols <-
  names(intensities)[!(names(intensities) %in% compound.failure.frame$compound_id)]
dim(intensities)
intensities <- intensities[,keep.cols]
dim(intensities)
write.table(intensities,
            "output/compound_intensities_clean.txt",
            sep="\t", quote=F, row.names=F, col.names=T)

```

## Session Info

```{r}
sessionInfo()
```

